Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-04-23 paper/ms.tex
Small changes to abstract

  Elongated bar-like features are ubiquitous, occurring at the centers of
  approximately two-thirds of spiral disk galaxies \cite{2000AJ....119..536E,
  2007ApJ...657..790M}. Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute between galactic components over long (Gyr)
  timescales in galaxies hosting a bar \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work has overwhelmingly
  provided the expectation that, due to the interaction between the bar and the halo, the bar pattern
  will slow its rotation over time \cite{1992ApJ...400...80H,
  2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
  2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
  2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. We have
  performed a simulation of an isolated galactic disk hosting a strong bar
  which includes a state-of-the-art model of the interstellar medium and a live dark matter halo. Here we
  show that in this simulation the bar pattern does not slow down over time,
  and instead remains at a stable, constant rate of rotation. This behavior
  has been observed in previous simulations but its explanation has remained
  elusive.\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark matter halo act in concert to stabilize the bar pattern and prevent it from slowing down or speeding up. This result
  naturally explains why nearly all observed bars rotate quickly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G} and is especially relevant for
  our understanding of how the Milky Way arrived at its present day state.

@@

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar. Thus, numerical work has
shown that the gas phase has an outsized influence on the
bar.\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}

>>>>

  Elongated bar-like features are ubiquitous, occurring at the centers of
  approximately two-thirds of spiral disk galaxies \cite{2000AJ....119..536E,
  2007ApJ...657..790M}. Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute between galactic components over long (Gyr)
  timescales in galaxies hosting a bar \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work has overwhelmingly
  provided the expectation that, due to the interaction between the bar and the dark matter halo, the bar pattern
  will slow its rotation over time \cite{1992ApJ...400...80H,
  2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
  2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
  2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. We have
  performed a simulation of a Milky Way-like galactic disk hosting a strong bar
  which includes a state-of-the-art model of the interstellar medium and a live dark matter halo. In this simulation the bar pattern does not slow down over time,
  and instead remains at a stable, constant rate of rotation. This behavior
  has been observed in previous simulations but its explanation has remained
  elusive.\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark matter halo act in concert to stabilize the bar pattern and prevent it from slowing down or speeding up. This result
  naturally explains why nearly all observed bars rotate quickly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G} and is especially relevant for
  our understanding of how the Milky Way arrived at its present day state.

@@

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar.\cite{2011MNRAS.415.1027H} Thus, numerical work has
shown that the gas phase has an outsized influence on the
bar.\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}
